Towards an Ecological Trait-data Standard

### 2019-03-28 manuscript/manuscript.Rmd
final edits for resubmission of manuscript

1. Trait-based approaches are widespread throughout ecological research as they offer great potential to achieve a general understanding of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data providers. 
4. The use of a standard vocabulary allows trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our vocabulary may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the *Darwin Core Standard* for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe any kind of measurement and occurrence of organisms. The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details. 
A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12]. The terminologies provided by these standards are quite universal and cover most use cases of trait data. 
<!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
Both DwC and ABCD are ratified standards of the Biodiversity Information Standards (TDWG, www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.

@@

Nonetheless, we are aware of the diverse use-cases of trait data that might not yet be covered by the current version of the vocabulary. The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research. 
For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org). This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
We invite communities of all trait-based research fields to discuss, revise and submit terms and extensions of the vocabulary (coordinated via Github Issues at https://github.com/EcologicalTraitData/ETS/issues).
The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets.  a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers. Additional identifiers relate each row to other layers of information on d) the taxon resolution, the specimen (i.e., occurrence) or the origin or confidence on the reported measurement or fact.}

@@

- The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.  <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)-->
- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g., variation in measurement method) or aggregated values (e.g., statistical method), as well as the source of reported facts (e.g., literature source or expert reference). <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)-->
- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual specimens, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time. <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)-->

<!--Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content. A unique identifier links to these other datasheets, encoding single measurements or reported facts (`measurementID`), individual occurrences of a species (`occurrenceID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`). 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].

@@

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset , e.g., by entering "Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2019, Ecological Traitdata Standard Vocabulary v0.10, DOI:  10.5281/zenodo.1441132, URL: https://terminologies.gfbio.org/terms/ets/pages/" in the metadata field `conformsTo`. 
Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their individual URIs.

@@

To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, that will be of use for the exchange and handling of trait data in the context of the semantic web.

@@

Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary. We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.

>>>>

1. Trait-based approaches are widespread throughout ecological research as they offer great potential to achieve a general understanding of a wide range of ecological and evolutionary mechanisms. Accordingly, a wealth of trait data is available for many organism groups, but this data is underexploited due to a lack of standardisation and heterogeneity in data formats and definitions.
2. We review current initiatives and structures developed for standardising trait data and discuss the importance of standardisation for trait data hosted in distributed open-access repositories. 
3. In order to facilitate the standardisation and harmonisation of distributed trait datasets by data providers and data users, we propose a standardised vocabulary that can be used for storing and sharing ecological trait data. We discuss potential incentives and challenges for the wide adoption of such a standard by data providers. 
4. The use of a standard vocabulary allows for trait datasets from heterogeneous sources to be aggregated more easily into compilations and facilitates the creation of interfaces between software tools for trait-data handling and analysis. By aiding decentralised trait-data standardisation, our vocabulary may ease data integration and use of trait data for a broader ecological research community and enable global syntheses across a wide range of taxa and ecosystems.

@@

Standard terms for the formal description of the common concepts of biodiversity knowledge have been provided in the schema for biological collection records [Access to Biological Collection Data, ABCD; @holetschek12] or the *Darwin Core Standard* for biodiversity data [DwC; @wieczorek12]. 
These frameworks specify terms and classes to describe any kind of measurement and occurrence of organisms. The terms may be used for  defining columns in data tables that contain measurement values, units and categorical levels, taxon names, variables such as sex or life stage, information of time and date of observation, and methodological details. 
A suite of terminology extensions links to and expands the capacities of DwC [@wieczorek12]. Of particular importance for trait data is the 'MeasurementOrFact' extension, which typically would be used in database management and bioinformatics to structure trait observations [@parr16]. 
<!--For example, the Global Biodiversity Information Facility (GBIF, www.gbif.org) uses DwC terms. -->
Both DwC and ABCD are ratified standards of the Biodiversity Information Standards (TDWG, www.tdwg.org) which is a global network to support the development and wide adoption of exchange standards for biodiversity data.

@@

Nonetheless, we are aware of the diverse use-cases of trait data that might not yet be covered by the current version of the vocabulary. The version presented here is a mere starting point of a community effort towards a consolidated and comprehensive Ecological Trait-data Standard Vocabulary, as a key resource for trait-data standardisation in ecological research. 
For future development of the vocabulary, we will engage with a broader community of trait researchers, in particular via the Open Traits Network (http://opentraits.org), and ensure full compatibility with other initiatives of biodiversity data standardisation by collaborating with Biodiversity Information Standards TDWG (Taxonomic Databases Working Group, http://www.tdwg.org). This will also link our initiative to other trait-based research fields, like biomedical and agricultural research.
We invite communities of all trait-based research fields to discuss, revise and submit terms and extensions of the vocabulary (coordinated via Github Issues at https://github.com/EcologicalTraitData/ETS/issues).
The standard vocabulary will be released in subsequent versions and published as a stable reference on the GFBio Terminology Service.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\textwidth, angle=0]{figures/fig2.png}
\caption{Formats used for trait datasets.  a) taxon-level trait data compiled from literature or aggregated from measurements are often published as a compiled species$\times$traits wide-table; b) observation long-tables are a well defined and tidy data format, reporting one single measurement per row and relating it to a standard trait definition and accepted taxon name; c) additional columns may provide original names for maintaining author-side continuity, identifiers reference to taxa and trait concepts via unambiguous URI pointers. Additional identifiers relate each row to other layers of information on d) the taxon resolution, the individual organism (i.e. occurrence), or the origin or confidence on the reported measurement or fact.}

@@

- The `Taxon` extension provides further terms for specifying the taxonomic resolution of the observation and to ensure the correct reference in case of synonyms and homonyms.  <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-taxon)-->
- The `MeasurementOrFact` extension  provides terms to describe information at the level of single measurements or reported facts, such as the original literature reference for the reported value, the method of measurement or statistical method of aggregation. It provides important information that allows for the tracking of potential sources of noise or bias in measured data (e.g., variation in measurement method) or aggregated values (e.g., statistical method), as well as the source of reported facts (e.g., literature source or expert reference). <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-measurement-or-fact)-->
- The `Occurrence` extension contains vocabulary to describe information on the observation context of individual organisms, such as sex, life stage or age. This also includes the method of sampling and preservation, as well as the date and geographical location, which provide an important resource to analyse trait variation due to differences in space and time. <!--(https://terminologies.gfbio.org/terms/ets/pages/#extension-occurrence)-->

<!--Many terms of these extensions refine or copy terms of the DwC and their own [Taxon](http://rs.tdwg.org/dwc/terms/index.htm#Taxon),  [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions and EOL TraitBank's use of those terms (http://eol.org/info/structured_data_archives).
 If no taxonID has been provided, in order to be unambiguous about the taxon referred to by the trait observation, it is highly recommended to keep an additional reference to higher taxonomic hierarchies (using columns `kingdom`, `order` etc.) to avoid taxonomic homonyms, and provide the authority (???) and key of reference (???).  
Furthermore, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for traits recorded at the family or genus level. This information can then be used for trait inference at higher taxonomic resolution. <!-- using hierarchical probabilistic matrix factorisation [@schrodt15]. -->
These additional layers of information can either be added as extra columns to the core dataset or kept in separate data sheets, thus avoiding redundancy and duplication of content. A unique identifier links to these other datasheets, encoding single measurements or reported facts (`measurementID`), individual organisms of a species (`occurrenceID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).

The concept of 'occurrence' is prone to cause confusion. By definition of DwC it is "An existence of an Organism at a particular place at a particular time", thus any individual observed twice would have two distinct 'occurrences'. In most ecological contexts, sampling of an individual is only performed once, resulting in any occurrence being semantically identical with the individual organism (i.e., the DwC term 'organism').
Also, as 'occurrence' is strictly defined by a date-time event, it may mean the same as the common-sense concept of 'observation'. 
Some data-types may directly refer to existing global identifiers for occurrence IDs, e.g., a GBIF URI or a stable identifier references the precise specimen from which the measurement was taken [@groom17; @guntsch17].
As occurrence is the narrower term and the key concept for linking an individual organism to a place and time in DwC, and since it is indeed relevant to distinguish between multiple 'occurrences' of the same organism in some trait-based research applications, the ETS sticks to this terminology.

@@

In order to ensure traceability, the metadata of any dataset that employs this vocabulary should refer to the specific online version that was used to build the dataset , e.g., by entering "Schneider, F.D., Jochum, M., Le Provost, G., Penone, C., Ostrowski, A. and Simons, N.K., 2019, Ecological Traitdata Standard Vocabulary v0.10, DOI:  10.5281/zenodo.2605377, URL: https://terminologies.gfbio.org/terms/ets/pages/" in the metadata field `conformsTo`. 
Wherever referring to individual terms of the vocabulary in publications or metadata, this should be done via their individual URIs.

@@

To serve the demand for the standardisation and harmonisation of ecological trait data which has arisen from a growing number of distributed datasets of different research contexts, we propose a versatile vocabulary for the publication of new datasets, for the creation of data compilations, and for the exchange and handling of trait data in the context of the semantic web.

@@

Thanks to all respondents to an internal online survey on trait data for the Biodiversity Exploratories project and to Diana Bowler, Klaus Birkhofer, Runa Boeddinghaus, Markus Fischer, Jens Kattge (and the TRY Steering Commitee), Felicitas Löffler, Catrin Westphal and two anonymous reviewers for comments on the manuscript drafts and pre-print, as well as the ecological trait-data standard vocabulary. We are grateful to the organisers and participants of the Open Traits workshop in New Orleans, USA, in August 2018.
